---
title: "Conducting Correlations in R"
subtitle: "A friendly guide into R"
author: "Israel Arevalo"
date: "`r Sys.Date()`"
output: 
  github_document:
    toc: true
    
---


Correlation is a statistical measure that can be used to determine the strength and direction of the relationship between two variables. In R, the `cor()` function can be used to calculate correlation coefficients. In this tutorial, we will use simulated educational data to demonstrate how to conduct correlations in R.

# Data Preparation

First, we will need to load in the necessary packages and create the simulated data set. For this tutorial, we will create a data set that contains information on students' math scores and reading scores.

```{r}

set.seed(123)  # for reproducibility

# Loading libraries
library(ggplot2)

# create the data set
data <- data.frame(student_id = 1:30,
                   math_score = rnorm(30, mean = 70, sd = 10),
                   reading_score = rnorm(30, mean = 75, sd = 12))

```


Next, we will take a look at the structure of the data set using the `str()` function.

```{r}

str(data)

```

# Conducting Correlations

Now that we have the data loaded, we can calculate the correlation coefficient between the `math_score` and `reading_score` variables. The `cor.test()` function can be used to calculate the Pearson's correlation coefficient, which is the most commonly used correlation coefficient.

```{r}

cor.test(data$math_score, data$reading_score)


```

The correlation coefficient output, `-0.1574511`, is a number between -1 and 1 that indicates the strength and direction of the relationship between the two variables, in this case `math_score` and `reading_score`.

A positive coefficient indicates a positive correlation, meaning that as one variable increases, the other variable also tends to increase. A negative coefficient indicates a negative correlation, meaning that as one variable increases, the other variable tends to decrease. A coefficient of 0 indicates no correlation.

In this case, the correlation coefficient is negative, which suggests that there is a small negative correlation between the `math_score` and `reading_score` variables. This means that as math scores increase, reading scores tend to decrease slightly. The correlation coefficient of `-0.1574511` is quite close to 0, which indicates that the relationship between the two variables is weak. This means that although there is a weak negative correlation between math scores and reading scores, it is not very strong.

In order to test whether the correlation coefficient obtained is statistically significant, we can use a hypothesis test. The null hypothesis is that there is no correlation between the two variables, and the alternative hypothesis is that there is a correlation.

The test statistic used to test the significance of the correlation coefficient is the t-value. The t-value is calculated by dividing the correlation coefficient by its standard error. The standard error of the correlation coefficient is calculated using a specific formula that takes into account the sample size, the standard deviation of the two variables and the correlation coefficient.

To test the significance of the correlation coefficient, we use the t-value to calculate a p-value. The p-value represents the probability of observing a t-value as large or larger than the one computed from the data, assuming that the null hypothesis is true. A low p-value (typically less than 0.05) indicates that the correlation coefficient is statistically significant, and the null hypothesis can be rejected.

We can see these values available on our table output. The t-value is a measure of how different the correlation coefficient is from zero. The larger the t-value, the more evidence we have that there is a correlation between the two variables. The t-value in this case is `-0.84368`.

The next value is the p-value, which represents the probability of observing a t-value as large or larger than the one computed from the data, assuming that the null hypothesis is true. The p-value in this case is `0.406` which is greater than 0.05 (commonly used significance level). Therefore, we fail to reject the null hypothesis that there is no correlation between the two variables.

The last value is the confidence interval, which is a range of values that is likely to contain the true correlation coefficient. The 95% confidence interval in this case is between `-0.4899292` and `0.2150148`. This means that if we were to conduct this test multiple times, 95% of the time the true correlation coefficient would fall within this range.


# Visualizing the Results

Finally, we can create a scatter plot to visualize the relationship between the two variables.

```{r}

ggplot(data, aes(x = math_score, y = reading_score)) +
  geom_smooth(method = "lm", color = "red") +
  ggtitle("Scatter plot of Math Scores vs Reading Scores")

```

This plot shows the relationship between the `math_score` and `reading_score` variables, and can help to further illustrate the strength and direction of the correlation. The line of best fit (in red) shows the direction of the correlation and the slope of the line represents the strength of the correlation.

In summary, this tutorial demonstrated how to conduct a correlation analysis using simulated educational data in R. The steps outlined in this tutorial can be applied to any data set and any set of variables. Correlation analysis is a useful tool for identifying the relationship between two variables, and it can provide insights into the extent to which two variables are related. Additionally, visualizing the results using the `ggplot2` package can help to further understand the relationship between the variables.

Please keep in mind that correlation does not imply causation, just because two variables are correlated does not mean that one variable causes the other. It's important to consider other factors that may be impacting the relationship between the variables.
